---
title: "T2D Data Prep"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

# Setup

load in libraries
 
```{r, message=FALSE}
library(curatedMetagenomicData)
library(SummarizedExperiment)
library(dplyr)
library(compositions)
```


Paths for saving and / or loading

```{r}
# downloaded data path
raw.data.path.CMD = "all_expdata_raw.rds"
processed.data.path.CMD = "QS_expdata_processed.rds"
```


# Initial exploration

what type of data is available for a particular study

```{r}
# curatedMetagenomicData("QinJ_2012.+")
```

Look at available data for a certain study and type

```{r}
# # tried 'relative_abundance' and it looks like standard summarised experiment object
# # tried 'pathway_abundance' and it seems good
# studyTostudy = "QinJ_2012.relative_abundance"
# expdat = curatedMetagenomicData(pattern = studyTostudy, dryrun = FALSE, counts = TRUE, rownames = "short") 
# 
# # the assay
# expdat.assay = assay(expdat[[1]])
# View(expdat.assay)
# dim(expdat.assay)
# 
# # feature metadata
# feature.metadata = as.data.frame(rowData(expdat[[1]]))
# View(feature.metadata)
# # feature metadata only available for relative abundance, not for pathway abundance
# 
# # sample metadata
# sample.metadata = as.data.frame(colData(expdat[[1]]))
# View(sample.metadata)
# View(sample.metadata[c("study_name", "subject_id", "study_condition", "disease")])
# table(sample.metadata$study_condition, sample.metadata$disease)
```

# Obtain data

Collate taxa and pathway data for the 6 CMD studies

Load in raw data if already downloaded from the repository and saved

```{r}
all_expdata_raw = readRDS(raw.data.path.CMD)

# Filter studies based on previous work?
all_study_names = names(all_expdata_raw)
OS_study_names = c("HMP_2019_t2d" ,"FengQ_2015", "KarlssonFH_2013", "LiJ_2014", "YuJ_2015")
FS_study_names = c("HMP_2019_t2d" , "YuJ_2015", "SankaranarayananK_2015")
# Or based on the study design?
QS_study_names = all_study_names[all_study_names!="HMP_2019_t2d"]

# FILTER AT THIS STEP
all_expdata_raw = all_expdata_raw[QS_study_names]

```

Or fetch from the repository

```{r, message=FALSE}

# study_names = c("HMP_2019_t2d", "FengQ_2015", "KarlssonFH_2013", "LiJ_2014", "YuJ_2015", "SankaranarayananK_2015")
# 
# all_expdata_raw = sapply(study_names, function(study_name){
# 
#   relative_abundance = curatedMetagenomicData(pattern = paste0(study_name,".relative_abundance"), dryrun = FALSE, counts = TRUE, rownames = "short")
#   pathway_abundance = curatedMetagenomicData(pattern = paste0(study_name,".pathway_abundance"), dryrun = FALSE, counts = TRUE, rownames = "short")
# 
#   expdat = list(
#     species = assay(relative_abundance[[1]]),
#     species_metadata = as.data.frame(rowData(relative_abundance[[1]])),
#     pathways = assay(pathway_abundance[[1]]),
#     sample_metadata = as.data.frame(colData(relative_abundance[[1]]))
#   )
# 
# }, simplify = FALSE, USE.NAMES = TRUE)
# 
# 
# saveRDS(object = all_expdata_raw, file = raw.data.path.CMD)
```


# Pre-process data

Load in saved file of pre-processed data if already done

```{r}
# all_expdata = readRDS(processed.data.path.CMD)
# study_names = names(all_expdata)
```

or carry out the preprocessing and save

## Pre-process by study

first check the sample metadata to see how to filter

```{r}
# sample_conditions = lapply(all_expdata_raw, function(expdata){
#   table(expdata$sample_metadata$study_condition, expdata$sample_metadata$disease)
# })
# sample_conditions[[6]]
```


### Fix up pathways data

```{r}
all_expdata <- lapply(all_expdata_raw, function(expdata){
  
  # extract matrix and turn into long format data frame
  pathways_df <- as.data.frame(
    t(expdata$pathways)
  )
  pathways_df$sample_id <- rownames(pathways_df)
  
  pathways_df <- reshape(
    data = pathways_df,
    direction = "long",
    idvar = c("sample_id"),
    varying = list(colnames(pathways_df)[-ncol(pathways_df)]),
    v.names = "value",
    timevar = "feature_id",
    times = colnames(pathways_df)[-ncol(pathways_df)]
  )
  
  # extract shorter names
  pathways_df$feature_name_short_position <- unlist(regexpr(":", pathways_df$feature_id))-1
  pathways_df$feature_name_long_position <- unlist(regexpr("\\|", pathways_df$feature_id))-1
  
  pathways_df$feature_name_long_beg <- ((pathways_df$feature_name_short_position < 0) * 1) + ((pathways_df$feature_name_short_position >= 0) * (pathways_df$feature_name_short_position + 3)) 
  pathways_df$feature_name_long_end <- ((pathways_df$feature_name_long_position < 0) * nchar(pathways_df$feature_id)) + ((pathways_df$feature_name_long_position >= 0) * pathways_df$feature_name_long_position)
  pathways_df$feature_name_short_end <- ((pathways_df$feature_name_short_position < 0) * pathways_df$feature_name_long_end) + ((pathways_df$feature_name_short_position >= 0) * pathways_df$feature_name_short_position) 
  
  pathways_df$feature_name_short <- substr(pathways_df$feature_id, 1, pathways_df$feature_name_short_end)
  pathways_df$feature_name_long <- substr(pathways_df$feature_id, pathways_df$feature_name_long_beg, pathways_df$feature_name_long_end)
  
  # filter out unknown pathways
  pathways_df <- pathways_df[!is.element(pathways_df$feature_name_short,c('UNMAPPED', 'UNINTEGRATED')),]
  
  # aggregate for metadata
  pathways_metadata <- pathways_df %>%
    distinct(feature_name_short, feature_name_long) %>%
    as.data.frame()
  rownames(pathways_metadata) <- pathways_metadata$feature_name_short
  
  # aggregate to recreate the abundance matrix
  pathways_df <- pathways_df %>%
    group_by(sample_id, feature_name_short) %>%
    summarise(total_value = sum(value, na.rm = TRUE)) %>%
    as.data.frame()
  
  # reshape
  pathways_df <- reshape(
    data = pathways_df,
    direction = "wide",
    idvar = "feature_name_short",
    timevar = "sample_id"
  )
  
  # tidy up names
  rownames(pathways_df) <- pathways_df$feature_name_short
  pathways_df <- pathways_df[,-1]
  colnames(pathways_df) <- substr(colnames(pathways_df),13,nchar(colnames(pathways_df)))
  
  # return matrix with samples in the correct order
  pathways_mat <- as.matrix(pathways_df)
  pathways_mat <- pathways_mat[,rownames(expdata$sample_metadata)]
  
  # return list
  expdata = list(
    species = expdata$species,
    species_metadata = expdata$species_metadata,
    pathways = pathways_mat,
    pathways_metadata = pathways_metadata,
    sample_metadata = expdata$sample_metadata
  )
  
})

```

Here I am filtering by study and performing the CLR transformation before merging. 
After this step, features are removed if not observed in enough studies

### Filter by study

```{r}
filt_th <- 10
# features included if > filt_th% of samples have a val != 0 in a given study for at least 1 response group

all_expdata = lapply(all_expdata, function(expdata){
  
  # filter samples
  # only healthy or T2D samples
  samples_include = rownames(expdata$sample_metadata)[grepl("healthy",expdata$sample_metadata$disease) | grepl("T2D",expdata$sample_metadata$disease)]
  # EXTRA FILTERING TO REMOVE TUMOUR SAMPLES?
  # NO - we have decided to include cancer samples due to the presence of other comorbidities
  # if(processed.data.path.CMD == "QS_expdata_processed.rds"){
  #   samples_include = samples_include[is.element(samples_include,rownames(expdata$sample_metadata)[!is.element(expdata$sample_metadata$study_condition,c("adenoma","CRC"))])]
  # }
  # intersecting samples
  samples_include = samples_include[is.element(samples_include,colnames(expdata$species))]
  samples_include = samples_include[is.element(samples_include,colnames(expdata$pathways))]
  # filter samples
  sample_metadata = expdata$sample_metadata[samples_include,]
  # add in response variable
  sample_metadata$response = sample_metadata$disease
  sample_metadata$response[sample_metadata$response!="healthy"] = "T2D"
  species = expdata$species[,samples_include]
  pathways = expdata$pathways[,samples_include]
  
  # filter species and pathways by group
  samples_healthy = rownames(sample_metadata)[sample_metadata$response=="healthy"]
  samples_T2D = rownames(sample_metadata)[sample_metadata$response=="T2D"]
  
  # filter species
  # incl_healthy = ((rowSums(species[,samples_healthy], na.rm = TRUE)/sum(species[,samples_healthy], na.rm = TRUE))*100) > filt_th
  # incl_T2D = ((rowSums(species[,samples_T2D], na.rm = TRUE)/sum(species[,samples_T2D], na.rm = TRUE))*100) > filt_th
  incl_healthy = ((rowSums(species[,samples_healthy] != 0, na.rm = TRUE) / ncol(species[,samples_healthy])) * 100) >= filt_th
  incl_T2D = ((rowSums(species[,samples_T2D] != 0, na.rm = TRUE) / ncol(species[,samples_T2D])) * 100) >= filt_th
  species = species[incl_healthy | incl_T2D,]
  incl_var = apply(species, 1, var, na.rm=TRUE) > 0
  species = species[incl_var,]
  species_metadata = expdata$species_metadata[rownames(species),]
  
  # filter pathways
  # incl_healthy = ((rowSums(pathways[,samples_healthy], na.rm = TRUE)/sum(pathways[,samples_healthy], na.rm = TRUE))*100) > filt_th
  # incl_T2D = ((rowSums(pathways[,samples_T2D], na.rm = TRUE)/sum(pathways[,samples_T2D], na.rm = TRUE))*100) > filt_th
  incl_healthy = ((rowSums(pathways[,samples_healthy] != 0, na.rm = TRUE) / ncol(pathways[,samples_healthy])) * 100) >= filt_th
  incl_T2D = ((rowSums(pathways[,samples_T2D] != 0, na.rm = TRUE) / ncol(pathways[,samples_T2D])) * 100) >= filt_th
  pathways = pathways[incl_healthy | incl_T2D,]
  incl_var = apply(pathways, 1, var, na.rm=TRUE) > 0
  pathways = pathways[incl_var,]
  pathways_metadata = expdata$pathways_metadata[rownames(pathways),]
  
  # transform data with the clr transformation
  species = as.matrix(t(clr(x = t(species) + 1)))
  pathways = as.matrix(t(clr(x = t(pathways) + 1)))
  
  # return list
  expdata = list(
    species = species,
    species_metadata = species_metadata,
    pathways = pathways,
    pathways_metadata = pathways_metadata,
    sample_metadata = sample_metadata
  )
}
)

```


### Merge studies

identify features present in enough studies

```{r}
studies_obs_th <- length(all_expdata)
# species
species_incl = table(unlist(lapply(all_expdata, function(expdata){rownames(expdata$species)})))
species_incl = names(species_incl)[species_incl>=studies_obs_th]

# pathways
pathways_incl = table(unlist(lapply(all_expdata, function(expdata){rownames(expdata$pathways)})))
pathways_incl = names(pathways_incl)[pathways_incl>=studies_obs_th]
```

merge the data from all studies and only include features observed in enough studies

```{r}
# species
species = lapply(all_expdata, function(expdata){
  expdata.mat = expdata$species[is.element(rownames(expdata$species),species_incl),]
  expdata.mat.pad = matrix(NA, 
                           nrow = length(species_incl), ncol = ncol(expdata.mat),
                           dimnames = list(species_incl, colnames(expdata.mat)))
  
  expdata.mat.pad[rownames(expdata.mat),] = expdata.mat
  return(expdata.mat.pad)
  })
species = do.call(cbind,species)

# species metadata
species_metadata = lapply(all_expdata, function(expdata){
  metadata = expdata$species_metadata[is.element(rownames(expdata$species_metadata),species_incl),]
  metadata$species_id = rownames(metadata)
  return(metadata)
  })

species_metadata = do.call(rbind,species_metadata)
species_metadata = species_metadata[match(species_incl,species_metadata$species_id),]
rownames(species_metadata) = species_metadata$species_id

# pathways
pathways = lapply(all_expdata, function(expdata){
  expdata.mat = expdata$pathways[is.element(rownames(expdata$pathways),pathways_incl),]
  expdata.mat.pad = matrix(NA, 
                           nrow = length(pathways_incl), ncol = ncol(expdata.mat),
                           dimnames = list(pathways_incl, colnames(expdata.mat)))
  
  expdata.mat.pad[rownames(expdata.mat),] = expdata.mat
  return(expdata.mat.pad)
  })
pathways = do.call(cbind,pathways)

# pathways metadata
pathways_metadata = lapply(all_expdata, function(expdata){
  metadata = expdata$pathways_metadata[is.element(rownames(expdata$pathways_metadata),pathways_incl),]
  metadata$pathways_id = rownames(metadata)
  return(metadata)
  })

pathways_metadata = do.call(rbind,pathways_metadata)
pathways_metadata = pathways_metadata[match(pathways_incl,pathways_metadata$pathways_id),]
rownames(pathways_metadata) = pathways_metadata$pathways_id

# sample metadata
sample_metadata = lapply(all_expdata, function(expdata){
  metadata = expdata$sample_metadata[,c("study_name","subject_id","response")]
  metadata$sample_id = rownames(metadata)
  return(metadata)
  })
sample_metadata = do.call(rbind,sample_metadata)
rownames(sample_metadata) = sample_metadata$sample_id

```

check the merged data

```{r}
# check names
mean(rownames(sample_metadata) == colnames(pathways))
mean(rownames(sample_metadata) == colnames(species))
mean(rownames(species_metadata)==rownames(species))
mean(rownames(pathways_metadata)==rownames(pathways))
unique(sample_metadata$study_name)
# check na values
sum(rowMeans(is.na(species))==1)
sum(colMeans(is.na(species))==1)
sum(rowMeans(is.na(pathways))==1)
sum(colMeans(is.na(pathways))==1)
# check variances
mean(apply(species, 1, var, na.rm=TRUE) > 0)
mean(apply(species, 2, var, na.rm=TRUE) > 0)
mean(apply(pathways, 1, var, na.rm=TRUE) > 0)
mean(apply(pathways, 2, var, na.rm=TRUE) > 0)
```

filter out samples that have no variance

```{r}
species_var = colnames(species)[apply(species, 2, var, na.rm=TRUE) > 0]
pathways_var = colnames(pathways)[apply(pathways, 2, var, na.rm=TRUE) > 0]
var_incl = species_var[is.element(species_var,pathways_var)]
sample_metadata = sample_metadata[var_incl,]
species = species[,var_incl]
pathways = pathways[,var_incl]
# then rerun checks above
```

What are the dims?

```{r}
dim(species)
dim(pathways)
```

Looks good - now save data for later

# Save data

```{r}
all_expdata = list(
  species = species,
  species_metadata = species_metadata,
  pathways = pathways,
  pathways_metadata = pathways_metadata,
  sample_metadata = sample_metadata
)

saveRDS(object = all_expdata, file = processed.data.path.CMD)
```

